---
title: "UK Lung Deaths Data Analysis"
output: html_notebook
---

```{r setup, include=FALSE}
# Load necessary libraries
library(ggplot2)
library(forecast)
library(dplyr)
```

## Introduction

This notebook analyzes the built-in UK Lung Deaths dataset. The dataset consists of monthly deaths from bronchitis, emphysema, and asthma in the UK, categorized by sex (male and female).

## Loading the Data

```{r load-data}
# Load the data
data("ldeaths")
data("mdeaths")
data("fdeaths")

# Create a data frame
ldeaths_df <- data.frame(
  date = time(ldeaths),
  total_deaths = as.numeric(ldeaths),
  male_deaths = as.numeric(mdeaths),
  female_deaths = as.numeric(fdeaths)
)

# Display the first few rows of the dataset
head(ldeaths_df)
```

## Basic Exploration

```{r basic-exploration}
# Summary statistics
summary(ldeaths_df)

# Check for missing values
sum(is.na(ldeaths_df))
```

## Visualization

### Total Deaths Over Time

```{r total-deaths-plot}
ggplot(ldeaths_df, aes(x = date, y = total_deaths)) +
  geom_line(color = "blue") +
  labs(title = "Total Lung Deaths Over Time", x = "Year", y = "Total Deaths")
```

### Male and Female Deaths Over Time

```{r male-female-deaths-plot}
ggplot(ldeaths_df) +
  geom_line(aes(x = date, y = male_deaths, color = "Male Deaths")) +
  geom_line(aes(x = date, y = female_deaths, color = "Female Deaths")) +
  labs(title = "Male and Female Lung Deaths Over Time", x = "Year", y = "Deaths") +
  scale_color_manual("", breaks = c("Male Deaths", "Female Deaths"), values = c("blue", "red"))
```

### Seasonal Decomposition

```{r seasonal-decomposition}
# Decompose the time series
ldeaths_decomp <- decompose(ts(ldeaths, frequency = 12))
mdeaths_decomp <- decompose(ts(mdeaths, frequency = 12))
fdeaths_decomp <- decompose(ts(fdeaths, frequency = 12))

# Plot the decompositions
plot(ldeaths_decomp)
plot(mdeaths_decomp)
plot(fdeaths_decomp)
```

## Statistical Analysis

### Correlation Between Male and Female Deaths

```{r correlation-analysis}
# Calculate correlation
correlation <- cor(ldeaths_df$male_deaths, ldeaths_df$female_deaths)
correlation
```

### Seasonal Autoregressive Integrated Moving Average (SARIMA) Model

```{r sarima-model}
# Fit SARIMA model for total deaths
sarima_model <- auto.arima(ldeaths)
summary(sarima_model)

# Forecast next 12 months
forecast_sarima <- forecast(sarima_model, h = 12)
plot(forecast_sarima, main = "SARIMA Forecast for Total Deaths")
```

## Conclusion

This analysis provided an overview of the UK Lung Deaths dataset, including visualizations and a simple SARIMA model for forecasting future deaths. The dataset shows clear seasonal patterns and a strong correlation between male and female deaths due to lung diseases.
```

This corrected version should now run without issues. You can save this content back into your R Markdown file and run it again.